---
title: "Multivariate analysis Endure data"
author: "Dries Bonte"
date: "2024-10-10"
output: word_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Multivariate analysis on species abundances

The manuscript by Van De Walle *et al.*. received reviewer comments as part of the PCI-Ecology recommendation process (preprint: https://doi.org/10.1101/2024.06.24.600350). The JSDM was only performed on a selection of the 50 most common species and only incorporating presence/absence data as inflated zero abundances give rise to convergence problems. To maximise data-use, we therefore added a multivariate analysis based on abundance data of an enlarged dataset. Also for this analysis, we only worked with the most common species (>15 individuals in a total set of 583 samples).

##Data and package loading
```{r dataloading, echo=FALSE, message=FALSE, results=FALSE}
library(vegan)
library(ggplot2)
library(dplyr)
library(here)

species <- read.table(here("suppl_RDA_DBonte","species_abundances_wide.txt"), sep = '\t', header=TRUE)
env <- read.table(here("suppl_RDA_DBonte","data_env.txt"), sep='\t',  header=TRUE)

#remove sum row & column
lastrow<-nrow(species)
species<-species[-c(lastrow),]

lastcolumn<-ncol(species)
species<-species[,1:(lastcolumn-1)]

#set species columns according to threshold minimal_species_abundance
#15 is the minimum for avoiding redundant distance measures

minimal_species_abundance <- 15
species<-species[, which(colSums(species[,2:ncol(species)]) > minimal_species_abundance)]


joined_df <- merge(env, species, by = "sample_ID")

#remove potential NAs
complete.cases(joined_df) 
summary(joined_df)


#remove empty rows
ncol(joined_df)
somrij<-rowSums(joined_df[, 14:ncol(joined_df)-1])
#print(somrij)
joined_df<-joined_df[somrij>0,]


# Split combined data back into species data and environmental data
lastcolumn<-ncol(joined_df)
species_data_complete <- joined_df[, 13:lastcolumn] 

#only first 10rows env data are important
env_data <- joined_df[, 1:10]
#summary(species_data_complete)

#na.omit(species_data_complete)
#complete.cases(species_data_complete) 
```

## RDA analysis and visualisation

```{r RDA, echo=FALSE}
# Perform RDA
#rda_result <- rda(species_data_complete ~ urbanization + proportion_50 + mt_stat_50 + vitality + X + Y, data = env_data)


#without urbanisation since this showed to be a correlated parameter with cover of Marram grass at the transect level, not used in the JSDM as well for similar reasons
rda_result <- rda(species_data_complete ~ proportion_50 + mt_stat_50 + vitality + X + Y, data = env_data)



# First inspection of plot
plot(rda_result)





# Get RDA scores for pretty plotting
scores <- scores(rda_result, display = "sites")
env_scores <- scores(rda_result, display = "bp")

env_scores
scores

# Prepare data frames for plotting
rda_df <- data.frame(RDA1 = scores[, 1], RDA2 = scores[, 2], 
                     District = as.factor(env_data$district))

env_scores_df <- as.data.frame(env_scores)
env_scores_df$Variable <- c("Marram cover","Marram Moran's I","Marram vitality","Longitude","Latitude")

#select only data for continuos predictors
env_scores_df <- env_scores_df[1:6,]
env_scores_df
env_scores_df$RDA1 <- env_scores_df$RDA1*4
env_scores_df$RDA2 <- env_scores_df$RDA2*4
# Modify Variable column to show desired labels
#env_scores_df$Variable <- gsub("River", "", env_scores_df$Variable)
#env_scores_df$Variable <- gsub("Method", "", env_scores_df$Variable)

# Extract variation explained to annotate axis labels
variation_explained <- round(rda_result$CA$eig[1:2] / sum(rda_result$CA$eig) * 100, 2)
print(variation_explained)

# Create the combined plot
combined_plot <- ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey") +  # Add dashed grey line for y = 0
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey") +  # Add dashed grey line for x = 0
  geom_point(data = rda_df, aes(x = RDA1, y = RDA2, color = District), alpha = 0.6) +  # Set alpha for transparency and size for distance
  geom_point(data = env_scores_df, aes(x = RDA1, y = RDA2), color = "black", shape = 18, size = 4, alpha = 0.5) +  # Display centroids of variables 
  geom_text(data = env_scores_df, aes(x = RDA1, y = RDA2, label = Variable), size = 4, vjust = 1.5) +
  labs(x = paste0("RDA1 (", variation_explained[1], "%)"),  
       y = paste0("RDA2 (", variation_explained[2], "%)"),  
       color = "District") +
  #scale_shape_manual(values = c(16, 17, 18), labels = c("Barbierbeek", "Herk", "Berwinne")) +  
  #scale_color_manual(values = c("blue", "deeppink4")) +  #include colours for variable 'Method'
  #scale_size_continuous(name = "Distance", range = c(3, 8), breaks = c(1000, 10000)) +  # Adjust size values for better differentiation in symbol sizes
  coord_fixed(ratio = 1) +  # Ensure equal scaling for x and y axes
  xlim(-4, 4) +
  ylim(-4, 4) + 
  theme_minimal() +
  theme(panel.grid = element_blank(),
        panel.border = element_rect(color = "black", fill = NA, size = 0.5),
        axis.title = element_text(size = 17),
        axis.text = element_text(size = 15))

# Print the combined plot
print(combined_plot)

# Print RDA summary
summary(rda_result)

#exported as 800 800 
```

## Variance partitioning using permanova
We use a complex design with nested factors (transects in regions, local marram variables within transects) and follow a sequential SS (typeI) to ensure each term is fitted after the previous one. In such a way, permanova mimics to the best the random design implemented in the JSDM. For local variables, we chose to fit vitality before the spatial variables, but in the end it does not matter for the variance partitioning. Permutations within strata 'transect' are included but are kind of redundant with the nested design. 

PERMANOVA analysis demonstrated the overall importance of biogeography as the most important explaining variable of the community structure (<!--this and next results may be subject to minor randomization variations when re-run--> R²=12%; Pseudo-F=18.35; df=5, P<0.001), but also the strong transect-dependent variation within the latter regions (R²=18%; Pseudo-F=3.88; df=38, P<0.001).  Additional variation was explained by the vitality (R²=7 %, Pseudo-F=1.29, df=43, P<0.001), cover (R²=7 %, Pseudo-F=1.24, df=44, P<0.001) and clustering (R²=6 %, Pseudo-F=1.13,  df=43, P=0.003) of the marram grass tussocks. 
```{r permanova, echo=FALSE}
distbb=vegdist(species_data_complete, "bray")

#first a test to 
#test1<-adonis2(distbb ~ factor(env_data$district)+env_data$X+env_data$Y+env_data$proportion_50+env_data$mt_stat_50+env_data$vitality, strata=env_data$transect , permutations=999)
#test2<-adonis2(distbb ~ factor(env_data$district)+env_data$proportion_50+env_data$mt_stat_50+env_data$vitality, strata=env_data$transect, permutations=999)


permaov <- adonis2(distbb ~ factor(env_data$district)+factor(env_data$transect) %in% factor(env_data$district)+env_data$vitality %in% factor(env_data$transect) +env_data$proportion_50 %in% factor(env_data$transect) +env_data$mt_stat_50 %in% factor(env_data$transect) , strata=env_data$transect
        , permutations=999, by="terms")

permaov
```
